---
title: 'Paired Samples'
author: "MAT150 - Stats for Life Sciences"
date: '2023-05-01'
output: html_document
---

<style>
    #header1 {
        color: #336699;
    }
    #normal {color: black;}
    p{font-size:15px;}
    h1{font-weight: bold;}
    h2{color: #336699;}
    h3{color: #666666;}
    ul{font-size:14px;}
    ol{font-size:15px;}
</style>


 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### R Commands covered in this lesson {#header1}

`t.test`, `pt`, `if`, `print`, `mean`, `sd`, `sqrt`, `as.numeric`

### Paired Samples for Comparing Two Populations or Treatement means
In many cases we are interested in comparing means before and after a treatment or between dependent individuals.  For example, 

* Pre and Post-treatment
* Twins
* Husband and wife


#### Summary

1. Null Hypotheis,  $H_0: \mu_d = D$

2. Test statistics
\[
t = \frac{\bar{x}_d - D}{s / \sqrt{n}}
\]


3. Calulate p-value using $df = n-1$.  

```{md}
p = pt(t,df,lower.tail = F)   # Right tail test
p = pt(t,df,lower.tail = T)   # Left tail test
p = 2*pt(t,df,lower.tail = F) # Two tail, but t > 0
p = 2*pt(t,df,lower.tail = T) # Two tail, but t < 0
```

4. Reject $H_0$ if $p < \alpha$.  

---


 #### Example
 
```{r}
memory = read.table('https://raw.githubusercontent.com/jamesquinlan/Intro-Stats-MAT150/main/data/examples/ex11_06.txt', header = T)
```


```{r}
pre = memory$memory_test[memory$test == "pre"]
post = memory$memory_test[memory$test == "post"]
results = t.test(pre, post, 
                 alternative = "less", 
                 paired = T, 
                 conf.level = 0.95)
n = length(pre)
tcrit = as.numeric(results$statistic["t"])
```

#### Confidence Interval

```{r}
a = mean(pre - post) -  tcrit * sd(pre - post)/sqrt(n)
b = mean(pre - post) +  tcrit * sd(pre - post)/sqrt(n)
c(a,b)
```


#### Example

Ultrasound is often used as treatment for soft tissue injury.  A range of motion study conducted on knee extension.  The data is: 

```{r}
pre  = c(31,53,45,57,50,43,32)
post = c(32,59,46,64,49,45,40)
```

Given 90% confidence level, is treatment effective?

```{r}
t.test(pre, post, 
       paired = T, 
       alternative = "less", 
       conf.level = 0.9)
```


#### Exercise

Do men overstate their height in online dating profiles?  We have
$\bar{x}_d = 0.57$ (profile - actual), $s_d = 0.81$, and $n = 40$.  

The null hypothesis is: $H_0: \mu_d = 0$ and $H_1: \mu_d > 0$.  

```{r}
# Aggregate Data
xd = 0.57
s  = 0.81
n = 40

# Calculate test statistic
t = xd/(s/sqrt(n))

# Calculate p-value
p = pt(t,n-1, lower.tail = F)

# Make Decision
if(p < 0.05){
  print("Reject the Null")
  }else{
  print("Fail to reject the Null")
}
```


### References

Peck, R., & Devore, J. L. (2011). *Statistics: The Exploration & Analysis of Data*. Cengage Learning.
