---
title: 'Describing Distribution Values in a Population'
author: "MAT150 - Stats for Life Sciences"
date: '2022-09-29'
output: html_document
---

<style>
    #header1 {
        color: #336699;
    }
    #normal {color: black;}
    p{font-size:15px;}
    h1{font-weight: bold;}
    h2{color: #336699;}
    h3{color: #666666;}
    ul{font-size:14px;}
    ol{font-size:15px;}
</style>


 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### R Commands covered in this lesson {#header1}

`barplot`,`hist`,`plot`,`dnorm`,`seq`


### Introduction

Our overall **goal** is to infer (generalize) from a *sample* to a *population*. 
A **population** is the entire collection of individuals or objects about which
information is desired.  For example, considering performance in a university statistics
class, the population consists of all students at university that take statistics.
The *variable of interest* might be the time it takes to complete an exam.  

Variables can be **categorical** or **numerical**.  

---

> DEFINTITION A **variable** associates a value with each individual or object in a population.
A variable can be either **categorical** or **numerical**, depending on the possible values.

---


### Categorical Variables

Categorical variables have relatively few number of possible (whole number) values (2, 3, 4, ...).
For example, *residence status* for the population of students at a large state university.
The values are: *resident* and *nonresident*.  Other examples of categorical variables include:

* smoker/nonsmoker
* race/ethnicity
* gender
* month of year
* favorite ice cream flavor
* concussions
* credit rating (good, poor)
* religious affiliation

##### Example

Consider the residence status for the population of students at a state university.
Students are either a resident of the state or not (dichotomous).  Plot the distribution
using a bar chart (see Example 7.1).

<!-- https://statisticsglobe.com/barplot-in-r -->
<!-- https://www.canva.com/colors/color-palettes/ -->

```{r}
group = c("Resident", "Nonresident")
values = c(0.73, 0.27)

width = 1

x = barplot(values, 
        names.arg = group, 
        col = c("#A2C4E0", "#FDB750"), 
        beside=TRUE,
        xlim = c(0,5), 
        ylim = c(0,0.8))
```



---

### Numerical Variables


> DEFINITION Numerical variables can be either **discrete** or **continuous**.  
A **discrete** variable is one whose possible values are isolated points along a number line.
Examples of discrete variables include the number of tweets, number of defective items, and the number of people in a checkout line. 
Typically, whole numbers (1, 2, 3, ...).  **Continuous* variables can take any value within an interval.
Examples include height, weight, and, temperature.

--- 

```{r}
values = c(rep(0,52),rep(1,22),rep(2,13), rep(3,9), rep(4,3), rep(5,1))
hist(values, 
     breaks = c(-0.5, 0.5, 1.5, 2.5, 3.5, 4.5, 5.5), 
     col = '#A2C4E0', 
     main = 'Distribution of Values', 
     ylim = c(0,60), 
     freq = TRUE)

```


#### 



--- 

### Summary

When describing data, the center and the spread are two key features.   
There are many quantities to describe how spread the data is including the range, the variance, the standard deviation, and the interquartile range.  


--- 

> DEFINITION: The **mean** value of a numerical variable $X$, denoted by $\mu$, describes the center of the population distribution.  The **standard deviation**, denoted $\sigma$, describes variability in the population distribution.  When $\sigma \approx 0$, the values of $X$ tend to be close to the mean.  (little variability). If $\sigma$ is large, there is more variability in the population of $X$ values.   

---


```{r}
x = rnorm(1000, 6, 2)
y = rnorm(1000, 11, 2)
par(mfrow=c(1,2))  
hist(x)
hist(y)

#reset par() options
# dev.off()
```
```{r}
#Create a sequence of 100 equally spaced numbers between -4 and 4
x = seq(-4, 4, length=100)

#create a vector of values that shows the height of the probability distribution
#for each value in x
y = dnorm(x, 0, 1)

#plot x and y as a scatterplot with connected lines (type = "l") and add
#an x-axis with custom labels
plot(x,y, type = "l", lwd = 2, axes = FALSE, xlab = "", ylab = "")
axis(1, at = -4:4, labels = c("-4s", "-3s", "-2s", "-1s", "0", "1s", "2s", "3s", "4s"))
```


### References

Peck, R., & Devore, J. L. (2011). *Statistics: The Exploration & Analysis of Data*. Cengage Learning.
