---
title: 'Statistics and Sampling Variability'
author: "MAT150 - Stats for Life Sciences"
date: '2022-09-29'
output: html_document
---

<style>
    #header1 {
        color: #336699;
    }
    #normal {color: black;}
    p{font-size:15px;}
    h1{font-weight: bold;}
    h2{color: #336699;}
    h3{color: #666666;}
    ul{font-size:14px;}
    ol{font-size:15px;}
</style>


 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### R Commands covered in this lesson {#header1}

`mean`,`sample`, `read.table`, `hist`, `choose`, `for`

## Statistics

---

> DEFINITION: A **statistic** is a quantity computed from sample data.  The sample mean, $\bar{x}$, sample median, and the sample standard deviation are all examples of statistics.   The value of the statistic will vary depending on the sample.  This variability is called **sample variability**.

---


Let $n$ be the sample size.  The **sample mean** is defined as:
\[
\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i
\]


### Example 8.1

Twenty students book expenses are given in the data below.  We will take samples
of this data and examine it.  

```{r}
# Read data
books = read.table('https://tinyurl.com/bdes8td3', header=TRUE)
```


The population mean is: 
```{r}
# Calculate (population) mean of the Amount spent on books
mean(books$Amount)
```

Take a sample of 5 students and calculate its mean (use as estimate of population mean).
```{r}
# Sample of 5 expense amounts
x = sample(books$Amount, 5)

# (Sample) Mean of those
mean(x)
```

Now, take 100 different samples (of 5) and calculate the mean of each sample.  
```{r}
N = 1:100

# Create vector of zeros
X = rep(0, length(N))
for (s in N) {
  X[s] = mean(sample(books$Amount,5))
}

# Show distribution using a Histogram
hist(X, main = 'Distribution of Sample Mean', col="#83bae2")
```


Note: there are $\binom{20}{5}$ unique samples of 5.   
```{r}
choose(20,5)
```



For completeness, calculate the **median** and **standard deviation** of a random sample.  
```{r}
median(x)
sd(x)
```



---

> DEFINITION: The distribution that would be formed by considering the value of 
a sample statistics for every possible different sample of a given size from a population is called the **sampling distribution**.

---



### Exercises

**8.2** What is the difference between $\bar{x}$ and $\mu$? between $s$ and $\sigma$?



**8.3** For each of the following statements, identify the number that appears in boldface type as the value of either a population characteristic or a statistic:

  a. A department store reports that **84%** of all customers who use the store's credit plan pay their bills on time.

  b. A sample of 100 students at a large university had a mean age of **24.1** years.

  c. The Department of Motor Vehicles reports that **22%** of all vehicles registered in a particular state are imports.

  d. A hospital reports that based on the 10 most recent cases, the mean length of stay for surgical patients is **6.4** days.

  e. A consumer group, after testing 100 batteries of a certain brand, reported an average life of **63** hours of use.



---


### Summary

We would like to know the value of a population characteristic (i.e., the mean), *however*, it is not feasible (or reasonable) to obtain.  Instead, we *sample* (take a subset), then calculate the value on the sample.  If the sample is drawn randomly, we should expect (depending on sample size), that it is a reasonably good estimation to the actual value.  If we take **ALL** random samples of a certain size, calculate the statistic, and create a histogram, we will see the histogram is fairly symmetric with one hump (i.e., normal).  This histogram is a visualization of the *sample distribution*.  The sample distribution of a statistics (i.e., sample mean $\bar{x}$) provides important information about variation in the values of the statistic. 



---

### References

Peck, R., & Devore, J. L. (2011). *Statistics: The Exploration & Analysis of Data*. Cengage Learning.
