---
title: "Homework-2"
output: html_document
date: "2022-09-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Questions

**1.12** Classify as categorical or numerical.  If numerical, determine whether
it is discrete or continuous. 

a. Number of students in a class of 35 who turn in a term paper before the due date.

**Numerical/Discrete**

b. Gender of the next baby born at a particular hospital.

**Categorical**

c. Amount of fluid (in ounces) dispensed by a machine used to fill bottles with soda pop.

**Numerical/continuous**

d. Thickness of the gelatin coating of a vitamin E capsule.

**Numerical/continuous**

e. Birth order classification (only child, firstborn, middle child, lastborn) of
a math major.

**Categorical**

---

**1.13** Classify each of the following variables as categorical or numerical.

a. Brand of computer: **categorical**
b. State of birth: **categorical**
c. Price of a textbook: **numerical**
d. Concentration of a contaminant: **numerical**
e. Zip code: **categorical**
f. Actual weight of coffee: **numerical**

---

**1.15** For each numerical variable, state whether it is discrete or continuous.

a. Length of rattlesnake: **continuous**
b. Altitude: **continuous**
c. Distance from ...: **continuous**
d. Price per gallon of gas: **discrete** (choosing discrete since $3.91234 isn't possible). 


---


**1.18** The following relative frequency distribution summarizing student responses 
to the question: "If you could make your college choice over, would you still
choose to enroll at your current college?"

| Response       | Freq  |
|:--------------|:-----|
| Definitely Yes | 0.447 |
| Probably Yes   | 0.373 |
| Probably no    | 0.134 |
| Definitely no  | 0.046 |

```{r}
# Create data frame from table above
choice = data.frame(
  Response = c("Definitely Yes", "Probably Yes", "Probably No", "Definitely No"),
  Freq = c(0.447, 0.373, 0.134, 0.046)
)
```

Make barplot:
```{r}
barplot(choice$Freq, names.arg = choice$Response)
```

---